# Reproducible Research: Peer Assessment 1

## Load and preprocess the data

> Process/transform the data (if necessary) into a format suitable for your analysis

```{r echo = TRUE}
dataZip <- "activity.zip"
dataFile <- "activity.csv"

# Skip if we already have the CSV
if (!file.exists(dataFile)) {
  unzip(dataZip)
}

act <- read.csv(dataFile)
```

## What is mean total number of steps taken per day?

> For this part of the assignment, you can ignore the missing values in the dataset.
> Make a histogram of the total number of steps taken each day
> Calculate and report the mean and median total number of steps taken per day

```{r echo = TRUE}
# Create data table and filter out incomplete cases
library(data.table)
# Creating data.table to aid in aggregate calculation
dt <- data.table(act[complete.cases(act),])

# Get aggregate sum for steps by day
aggAct <- dt[,list(steps=sum(steps)), by = date]
# Translate factors to dates for charting purposes
aggAct$date <- as.Date(aggAct$date)

# Make a histogram of the total number of steps taken each day
library(ggplot2)
qplot(date, steps, data = aggAct, xlab = "Dates", ylab="Total step count", geom="histogram", stat="identity") 

# Calculate and report the mean and median total number of steps taken per day.
# Interpreting this as asking for the mean and median of the daily step total values.
mean(aggAct$steps)
median(aggAct$steps)
```

## What is the average daily activity pattern?

> Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

> Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo = TRUE}
# Aggreagate by interval
aggAct <- dt[,list(steps=mean(steps)), by = interval]

# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken,
# averaged across all days (y-axis)
plot(aggAct$interval, aggAct$steps, type = "l", xlab="Interval", ylab="Average number of steps") 

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
aggAct[with(aggAct, order(-steps)), interval][1]
```

## Imputing missing values

> Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

> Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

> Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

> Create a new dataset that is equal to the original dataset but with the missing data filled in.

> Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo = TRUE}
# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows 
# with NAs)
length(which(is.na(act$steps)))

# Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be 
# sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

# Create a new dataset that is equal to the original dataset but with the missing data filled in.
# Substituting rounded mean value for interval NA value is from
act$steps[is.na(act$steps)] <- ave(act$steps, 
                                   act$interval, 
                                   FUN = function(x) round(mean(x, na.rm = TRUE)))[c(which(is.na(act$steps)))]
length(which(is.na(act$steps)))

# Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of
# steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact
# of imputing missing data on the estimates of the total daily number of steps?

# Creating data.table to aid in aggregate calculation
dt <- data.table(act)

# Get aggregate sum for steps by day
aggAct <- dt[,list(steps=sum(steps)), by = date]
# Translate factors to dates for charting purposes
aggAct$date <- as.Date(aggAct$date)

qplot(date, steps, data = aggAct, xlab = "Dates", ylab="Total step count", geom="histogram", stat="identity") 

# Was 10766 without imputing
mean(aggAct$steps)
# Was 10765 without imputing
median(aggAct$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

> For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

> Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

> Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data: <graph>

> Your plot will look different from the one above because you will be using the activity monitor data. Note that the above plot was made using the lattice system but you can make the same version of the plot using any plotting system you choose.

```{r echo = TRUE}
# Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date
# is a weekday or weekend day.
library(data.table)
dt <- data.table(act[complete.cases(act),])
dt$date <- as.Date(dt$date)
dt$date.type <- ifelse(is.element(weekdays(dt$date), c("Saturday","Sunday")), "weekend", "weekday")
dt$date.type <- as.factor(dt$date.type)

aggAct <- dt[,list(steps=mean(steps)), by = list(interval, date.type)]

# Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average
# number of steps taken, averaged across all weekday days or weekend days (y-axis).

par(mfrow = c(2, 1))
weekdayAggAct <- aggAct[date.type == "weekday",]
weekdayAggAct <- weekdayAggAct[,list(steps=mean(steps)), by = interval]
plot(weekdayAggAct$interval, weekdayAggAct$steps, type="l", xlab="Interval", ylab="Number of steps", main="Weekdays") 
weekendAggAct <- aggAct[date.type == "weekend",]
weekendAggAct <- weekendAggAct[,list(steps=mean(steps)), by = interval]
plot(weekendAggAct$interval, weekendAggAct$steps, type="l", xlab="Interval", ylab="Number of steps", main="Weekends") 
title( "Step Count Comparison", outer = TRUE )
```